General purpose gene annotation using deep learning

### 2020-04-21 paper/main.tex
many small things

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 (\citealp{Howe19}) as well as all non-embargoed plant genomes of the JGI Phytosome 13 database (\citealp{Phytozome}). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 and 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of non-test genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also do not find a noticible difference in performance between those two groups.

We pre-processed and stored the genomic information by using Geenuff (\citealp{Denton19}). Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Supplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{tab:arrays}.

@@

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation (explanation why this is a sign of error?). As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to effectively blank out those bases during training. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Section \ref{sec:training_data}.

@@

Once a set of training genomes was selected, we split it up again into a {\it training and validation set}. This split of the training genomes was done to get a quicker sense of the generalization capabilities and was used primarily during training. We split of the validation set by selecting 20\% of the FASTA sequences above and below the N90 of each training genome, ensuring a proper distribution of large and short sequences in both sets and a split on the chromosome level. Evaluation of the annotations on all training and validation genomes was done regularly after a promising model candidate was found based on its validation set performance. Ultimately we only cared about the cross-species performance of a model.

We use metrics two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Chapter \ref{augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes. All individual class performances are reported in Supplementary Material (?).

@@

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{tab:arrays}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in a one hot encoding\\
	  & for (A, T, C, G), otherwise different, see suplementary .. ()\\
Output & Labels in a 4-dimensional one hot encoding \\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{}
\end{table}

@@

\subsection{Input Data Encoding}

@@

>>>>

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 (\citealp{Howe19}) as well as all non-embargoed plant genomes of the JGI Phytosome 13 database (\citealp{Phytozome}). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 and 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of non-test genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also did not find a noticible difference in performance between those two groups.

We pre-processed and stored the genomic information by using Geenuff (\citealp{Denton19}). Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. The encoding of the genomic sequence is done in line with the IUPAC nucleic acid notation and the functional annotation used as labels during training is transformed to a one hot encoding with the four classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table \ref{tab:arrays} for a more detailed description of the generated data formats.

@@

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation. As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to blank out those bases during training by using the sample weights described in Table \ref{tab:arrays}. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Section \ref{sec:training_data}.

@@

Once a set of training genomes was selected, we split it up again into a {\it training and validation set}. This split of the training genomes was done to get a quicker sense of the generalization capabilities and was used primarily during training. We split off the validation set by selecting 20\% of the FASTA sequences above and below the N90 of each training genome, ensuring a proper distribution of large and short sequences in both sets and a split on the chromosome level. Evaluation of the annotations on all training and validation genomes was done regularly after a promising model candidate was found based on its validation set performance. Ultimately we only cared about the cross-species performance on all validation genomes of a model.

We use two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Section \ref{sec:augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. (See Table \ref{tab:statistics_all} for the class distribution and Table \ref{tab:detailed_results} for a more in-depth report our model performance including the intergenic class). In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes.

@@

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{tab:arrays}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in the 4-dimensional IUPAC encoding\\
	  & (one hot encoding for non-ambiguous bases) \\
Output & Labels in a 4-dimensional one hot encoding representing \\
      & the classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}\\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{An intron inside a UTR region is encoded as intron and can be inferred from the prediction order.}
\end{table}

@@

@@

\subsection{Detailed model performance}

\begin{table}[!t]
\processtable{Class weights\label{tab:detailed_results}} {
\begin{tabular}{@{}lll@{}}
\toprule Class & Animal Value & Plant Value \\
\midrule
Intergenic & 0.7 & 0.3 \\
UTR & 1.6 & 1.0 \\
CDS & 1.2 & 0.9 \\
Intron & 1.2 & 0.3 \\
\botrule
\end{tabular}}{}
\end{table}
